---
title: "Basic Workflow with polylinkR"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Basic Workflow with polylinkR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE
)
```

```{r setup, message=FALSE}
library(polylinkR)
library(data.table)
library(ggplot2)
```

## Introduction

**polylinkR** performs gene-based pathway enrichment analysis while explicitly accounting for linkage disequilibrium (LD) between adjacent loci. Unlike standard pathway enrichment tools that treat genes as independent, polylinkR preserves the innate linkage structure among genes across the genome. This reduces false positives that arise when genes in the same pathway are physically close and share correlated signals.

The package implements a novel circular permutation algorithm that:
1. Links all chromosomes into a single "circular" genome
2. Rotates this circular genome to create unique gene-to-score mappings
3. Preserves the correlation structure among genes
4. Generates accurate null distributions for pathway enrichment testing

## Prerequisites

### Required Packages

polylinkR requires the following R packages:

- `data.table` - Fast data manipulation
- `foreach` - Parallel iteration support  
- `doFuture` - Parallel backend
- `future` - Parallel processing framework
- `dqrng` - Fast random number generation
- `Rfast` - Fast statistical functions
- `mgcv` - Generalized additive models
- `distances` - Distance matrix calculations
- `tdigest` - Approximate quantile estimation
- `cli` - Command line interface formatting

These dependencies are automatically installed when you install polylinkR from CRAN or GitHub.

### Input Data Format

polylinkR requires three core input files:

1. **obj.info** - Gene-level information with columns:
   - `objID`: Unique gene identifier
   - `objStat`: Gene-level statistic (e.g., selection score, p-value)
   - `chr`, `startpos`, `endpos`: Genomic coordinates (optional but recommended)

2. **set.info** - Gene set information with columns:
   - `setID`: Unique pathway/set identifier
   - Additional optional metadata columns

3. **set.obj** - Gene-to-set mappings with columns:
   - `setID`: Set identifier (must match set.info)
   - `objID`: Gene identifier (must match obj.info)

Optional files include:
- **var.info** - SNP-level statistics for computing gene scores
- **rec.rate** - Recombination rates for converting physical to genetic distances

## Loading Data

### The Anatolia_EF_CLR Dataset

This vignette uses the `Anatolia_EF_CLR` dataset included with polylinkR. This dataset contains gene-level statistics from a population genetics study of the Anatolia population, with selection signals computed using Eigenstrat format data.

Let's examine the structure of the included data:

```{r load-data}
# Load the example datasets
data("Anatolia_EF_CLR")
data("PolyLinkR_SetInfo")
data("PolyLinkR_SetObj")
data("rr.dt")

# Examine the gene information
head(Anatolia_EF_CLR)
str(Anatolia_EF_CLR)

# Check the number of genes
cat("Number of genes:", nrow(Anatolia_EF_CLR), "\n")
cat("Number of chromosomes:", length(unique(Anatolia_EF_CLR$chr)), "\n")
```

The `Anatolia_EF_CLR` object contains:
- **objID**: Gene identifiers
- **objStat**: Gene selection statistics (higher values indicate stronger signals)
- **objName**: Gene symbols (e.g., A1BG, A2M)
- **chr**: Chromosome numbers
- **startpos/endpos**: Genomic positions in base pairs
- **GeneLength**: Gene length in base pairs

Let's also examine the gene sets:

```{r examine-sets}
# Gene set information
head(PolyLinkR_SetInfo)
cat("Number of gene sets:", nrow(PolyLinkR_SetInfo), "\n")

# Gene-to-set mappings
head(PolyLinkR_SetObj)
cat("Number of gene-set associations:", nrow(PolyLinkR_SetObj), "\n")

# Check how many genes are in each set (sample)
set_counts <- table(PolyLinkR_SetObj$setID)
summary(as.vector(set_counts))
```

### Data Format Summary

The Eigenstrat format typically includes:
- **.snp** files: SNP positions and allele information
- **.geno** files: Genotype data encoded in CLR (Complete Linkage Representation)
- **.ind** files: Individual/sample information

For polylinkR, we work with derived gene-level statistics that summarize these data, where each gene has an associated selection signal (`objStat`) and genomic coordinates.

## Basic Analysis Walkthrough

### Step 1: Reading Data with `read_polylinkr_data()`

The `read_polylinkr_data()` function is the entry point for polylinkR analysis. It validates input files, performs quality checks, and creates a `plR` class object that subsequent functions operate on.

#### Parameter Explanations

Key parameters when reading data:

- **`input.path`**: Directory containing all input files (alternative: specify individual file paths)
- **`obj.info.path`**, **`set.info.path`**, **`set.obj.path`**: Individual file paths
- **`rec.rate.path`**: Optional recombination rate file for coordinate conversion
- **`min.set.n`**: Minimum gene set size (default: 2, must be ≥ 2)
- **`max.set.n`**: Maximum gene set size (default: Inf)
- **`set.merge`**: Threshold for merging similar gene sets (default: 0.95 = 95% similarity)
- **`rem.genes`**: Whether to remove genes with identical positions (default: FALSE)

Since we have data loaded in R already, let's save it to temporary files and demonstrate the reading process:

```{r prepare-files}
# Create temporary directory for demonstration
temp_dir <- tempdir()

# Write data to files
data.table::fwrite(as.data.table(Anatolia_EF_CLR), 
                   file.path(temp_dir, "obj.info.tsv"), sep = "\t")
data.table::fwrite(as.data.table(PolyLinkR_SetInfo), 
                   file.path(temp_dir, "set.info.tsv"), sep = "\t")
data.table::fwrite(as.data.table(PolyLinkR_SetObj), 
                   file.path(temp_dir, "set.obj.tsv"), sep = "\t")
data.table::fwrite(as.data.table(rr.dt), 
                   file.path(temp_dir, "rec.rate.tsv"), sep = "\t")

# List the files
cat("Files prepared in:", temp_dir, "\n")
list.files(temp_dir, pattern = "\\.tsv$")
```

Now read the data into polylinkR:

```{r read-data, message=TRUE}
# Read the prepared data
plr_obj <- read_polylinkr_data(
  input.path = temp_dir,
  min.set.n = 5,      # Only include gene sets with 5+ genes
  set.merge = 0.95,   # Merge gene sets with ≥95% overlap
  verbose = TRUE
)

# Check the result
class(plr_obj)
print(plr_obj)
```

The `read_polylinkr_data()` function returns a `plR` S3 object containing three data.tables:
- **obj.info**: Gene information with validated scores
- **set.info**: Gene set information with sizes
- **set.obj**: Gene-to-set mappings

### Step 2: Permutation Testing with `permute_polylinkr_data()`

The `permute_polylinkr_data()` function performs gene set enrichment testing using a permutation procedure that:
1. Creates null distributions by permuting gene scores while preserving linkage structure
2. Fits Generalized Pareto Distributions (GPD) to the tails for accurate small p-values
3. Computes enrichment p-values for each gene set

#### Parameter Explanations

Key parameters for permutation:

- **`n_permutations`**: Number of permutations (default: 500,000). More permutations = more accurate p-values but slower. Minimum: 50,000.
- **`n_bootstraps`**: Number of bootstrap replicates for GPD fitting (default: 30). Higher = more stable but slower.
- **`alternative`**: Test direction - `"upper"` for enrichment (default) or `"lower"` for depletion
- **`gpd.cutoff`**: Tail probability threshold for GPD fitting (default: 500/n_permutations)
- **`seed`**: Random seed for reproducibility
- **`n.cores`**: Number of cores for parallel processing

For this demonstration, we'll use a smaller number of permutations for speed:

```{r permute-data, message=TRUE}
# Run permutation analysis with reduced permutations for vignette speed
plr_perm <- permute_polylinkr_data(
  plr_input = plr_obj,
  n_permutations = 50000,      # Use 50k for demonstration (use 500k+ for real analysis)
  n_bootstraps = 10,         # Reduced bootstrap replicates
  seed = 42,           # For reproducibility
  n.cores = 1,         # Single core for vignette
  verbose = TRUE
)

# Check the results
print(plr_perm)
```

The `permute_polylinkr_data()` function adds new columns to `set.info`:
- **setScore.std**: Standardized gene set scores
- **setScore.std.p**: P-values for enrichment

Let's examine the significant results:

```{r examine-permute-results}
# Extract set info with p-values
set_results <- plr_perm$set.info

# Show top significant sets
head(set_results[order(setScore.std.p)][, .(setID, setN, setScore.std, setScore.std.p)])

# Summary of p-values
summary(set_results$setScore.std.p)
```

### Step 3: Rescaling for Autocorrelation with `rescale_polylinkr_data()`

The `rescale_polylinkr_data()` function adjusts gene set scores to account for genetic autocorrelation - the tendency for nearby genes to have correlated statistics due to LD. This step is crucial for accurate p-value estimation.

#### Parameter Explanations

Key parameters for rescaling:

- **`rescale`**: Whether to perform rescaling (default: TRUE). Set FALSE to only estimate autocorrelation.
- **`fast`**: Use fast analytical rescaling (default: TRUE) vs. slower empirical rescaling
- **`cgm.range`**: Maximum inter-gene distance for autocovariance estimation (default: 2 Mbp or 2 cM)
- **`cgm.bin`**: Minimum gene pairs per lag bin (default: 30)
- **`min.rho`**: Minimum correlation threshold (default: 1e-5)

```{r rescale-data, message=TRUE}
# Rescale for genetic autocorrelation
plr_rescale <- rescale_polylinkr_data(
  plr_input = plr_perm,
  rescale = TRUE,
  fast = TRUE,         # Use fast analytical method
  verbose = TRUE
)

# Check results
print(plr_rescale)
```

The `rescale_polylinkr_data()` function adds:
- **setScore.rs**: Rescaled gene set scores (accounting for autocorrelation)
- **setScore.rs.p**: P-values for rescaled scores

Compare standard vs. rescaled scores:

```{r compare-rescale}
# Compare standard and rescaled results
comparison <- plr_rescale$set.info[, .(setID, setN, 
                                       setScore.std, setScore.std.p,
                                       setScore.rs, setScore.rs.p)]
head(comparison[order(setScore.rs.p)])
```

### Step 4: Pruning and FDR Correction with `prune_polylinkr_data()`

The `prune_polylinkr_data()` function performs multiple testing correction while accounting for shared genes between pathways. Gene sets often overlap, and this step ensures FDR control accounts for these dependencies.

#### Parameter Explanations

Key parameters for pruning:

- **`n.fdr`**: Number of permuted datasets for FDR estimation (default: 300). Must be divisible by 100.
- **`est.pi0`**: Whether to estimate proportion of true nulls (default: TRUE)
- **`tolerance`**: Convergence tolerance for pi0 estimation (default: 1e-3)

```{r prune-data, message=TRUE, eval=FALSE}
# Note: This step is computationally intensive and skipped in vignette
# Run it for your actual analyses

plr_final <- prune_polylinkr_data(
  plr_input = plr_rescale,
  n.fdr = 300,         # Number of FDR replicates
  est.pi0 = TRUE,      # Estimate proportion of true nulls
  verbose = TRUE
)

# Results include:
# - setScore.pr: Pruned gene set scores
# - setScore.pr.p: P-values for pruned scores  
# - setScore.pr.q: FDR-corrected q-values
```

## Result Visualization

### Manhattan-style Plot of -log10(p-values)

A common way to visualize pathway enrichment results is a Manhattan-style plot showing -log10(p-values) across gene sets:

```{r manhattan-plot}
# Prepare data for plotting
plot_data <- plr_rescale$set.info[, .(setID, setN, setScore.std.p, setScore.rs.p)]
plot_data[, log_p_std := -log10(setScore.std.p)]
plot_data[, log_p_rs := -log10(setScore.rs.p)]

# Create significance threshold line
threshold <- -log10(0.05 / nrow(plot_data))  # Bonferroni

# Plot
p <- ggplot(plot_data, aes(x = seq_along(setID), y = log_p_rs)) +
  geom_point(aes(color = log_p_rs > 5), alpha = 0.6, size = 2) +
  geom_hline(yintercept = threshold, linetype = "dashed", color = "red") +
  geom_hline(yintercept = -log10(0.05), linetype = "dotted", color = "blue") +
  scale_color_manual(values = c("grey50", "red"), 
                     labels = c("Not significant", "Suggestive")) +
  labs(
    title = "Gene Set Enrichment Results",
    subtitle = "polylinkR Analysis of Anatolia_EF_CLR Dataset",
    x = "Gene Set Index",
    y = expression(-log[10](p-value)),
    color = "Significance"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )

print(p)
```

### Volcano Plot: Effect Size vs. P-value

Another useful visualization is a volcano plot showing the relationship between effect size (gene set score) and statistical significance:

```{r volcano-plot}
# Prepare volcano plot data
volcano_data <- plr_rescale$set.info[, .(setID, setN, 
                                          setScore.rs, setScore.rs.p,
                                          setScore.std)]
volcano_data[, log_p := -log10(setScore.rs.p)]
volcano_data[, significance := ifelse(setScore.rs.p < 0.05 / nrow(volcano_data), 
                                    "Significant", 
                                    ifelse(setScore.rs.p < 0.05, 
                                           "Suggestive", "Not significant"))]

# Plot
p2 <- ggplot(volcano_data, aes(x = setScore.rs, y = log_p)) +
  geom_point(aes(color = significance, size = setN), alpha = 0.6) +
  scale_color_manual(values = c("Not significant" = "grey50", 
                               "Suggestive" = "orange",
                               "Significant" = "red")) +
  scale_size_continuous(range = c(1, 4)) +
  geom_hline(yintercept = -log10(0.05), linetype = "dotted") +
  labs(
    title = "Volcano Plot: Gene Set Enrichment",
    x = "Rescaled Gene Set Score",
    y = expression(-log[10](p-value)),
    color = "Significance",
    size = "Gene Set Size"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

print(p2)
```

### Gene Set Size Distribution

It's also useful to visualize the distribution of gene set sizes:

```{r set-size-plot}
# Plot gene set size distribution
p3 <- ggplot(plr_rescale$set.info, aes(x = setN)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7, color = "white") +
  scale_x_log10() +
  labs(
    title = "Distribution of Gene Set Sizes",
    x = "Number of Genes (log scale)",
    y = "Count"
  ) +
  theme_minimal()

print(p3)
```

## Saving Results

### Export to CSV/TSV Files

polylinkR results can be easily exported for further analysis or sharing:

```{r export-results}
# Export gene set results
results_to_export <- plr_rescale$set.info[, .(setID, setN, 
                                              setScore.std, setScore.std.p,
                                              setScore.rs, setScore.rs.p)]

# Add set names if available
if ("setName" %in% names(plr_rescale$set.info)) {
  results_to_export[, setName := plr_rescale$set.info$setName]
}

# Write to file
output_file <- file.path(temp_dir, "polylinkR_results.tsv")
data.table::fwrite(results_to_export, output_file, sep = "\t")

cat("Results saved to:", output_file, "\n")
cat("Number of gene sets:", nrow(results_to_export), "\n")
cat("Columns:", paste(names(results_to_export), collapse = ", "), "\n")
```

### Save Complete plR Object

For reproducibility and further analysis, save the complete plR object:

```{r save-object}
# Save the complete plR object
rds_file <- file.path(temp_dir, "polylinkR_results.rds")
saveRDS(plr_rescale, rds_file)

cat("Complete plR object saved to:", rds_file, "\n")

# Demonstrate loading
loaded_obj <- readRDS(rds_file)
cat("Loaded object class:", class(loaded_obj), "\n")
cat("Gene sets in loaded object:", nrow(loaded_obj$set.info), "\n")
```

### Accessing Attributes

plR objects contain rich metadata in attributes:

```{r access-attributes}
# Access attributes
attrs <- attributes(plr_rescale)
names(attrs)

# View function arguments used
attrs$plr.args$read.args$min.set.n
  attrs$plr.args$permute.args$n_permutations
attrs$plr.args$rescale.args$fast

# View data summaries
attrs$plr.data$read.data
```

## Complete Reproducible Example

Here is the complete workflow from start to finish:

```{r complete-example, message=FALSE}
# ============================================================
# COMPLETE POLYLINKR WORKFLOW - REPRODUCIBLE EXAMPLE
# ============================================================

# 1. LOAD PACKAGES
library(polylinkR)
library(data.table)

# 2. LOAD DATA
data("Anatolia_EF_CLR")
data("PolyLinkR_SetInfo")
data("PolyLinkR_SetObj")
data("rr.dt")

# 3. PREPARE INPUT FILES
temp_dir <- tempdir()
data.table::fwrite(as.data.table(Anatolia_EF_CLR), 
                   file.path(temp_dir, "obj.info.tsv"), sep = "\t")
data.table::fwrite(as.data.table(PolyLinkR_SetInfo), 
                   file.path(temp_dir, "set.info.tsv"), sep = "\t")
data.table::fwrite(as.data.table(PolyLinkR_SetObj), 
                   file.path(temp_dir, "set.obj.tsv"), sep = "\t")
data.table::fwrite(as.data.table(rr.dt), 
                   file.path(temp_dir, "rec.rate.tsv"), sep = "\t")

# 4. READ DATA
plr_obj <- read_polylinkr_data(
  input.path = temp_dir,
  min.set.n = 5,
  verbose = TRUE
)

# 5. PERMUTATION TESTING
plr_perm <- permute_polylinkr_data(
  plr_input = plr_obj,
  n_permutations = 50000,    # Use 500000+ for real analysis
  n_bootstraps = 10,
  seed = 42,
  verbose = TRUE
)

# 6. RESCALE FOR AUTOCORRELATION
plr_rescale <- rescale_polylinkr_data(
  plr_input = plr_perm,
  fast = TRUE,
  verbose = TRUE
)

# 7. EXTRACT AND SAVE RESULTS
results <- plr_rescale$set.info[, .(setID, setN, 
                                    setScore.std, setScore.std.p,
                                    setScore.rs, setScore.rs.p)]

# Add significance flags
results[, significant := setScore.rs.p < 0.05 / .N]  # Bonferroni
results[, suggestive := setScore.rs.p < 0.05]

# Summary of results
cat("\n=== ANALYSIS SUMMARY ===\n")
cat("Total gene sets analyzed:", nrow(results), "\n")
cat("Significant sets (Bonferroni):", sum(results$significant), "\n")
cat("Suggestive sets (p < 0.05):", sum(results$suggestive), "\n")

# Top 10 most significant sets
cat("\n=== TOP 10 SIGNIFICANT GENE SETS ===\n")
top_sets <- head(results[order(setScore.rs.p)], 10)
print(top_sets[, .(setID, setN, setScore.rs, setScore.rs.p, significant)])

# 8. CLEAN UP
unlink(temp_dir, recursive = TRUE)
cat("\n=== ANALYSIS COMPLETE ===\n")
```

## Tips and Common Pitfalls

### Tips for Successful Analysis

 1. **Start with smaller n_permutations**: For initial exploration, use n_permutations = 50,000. For publication-quality results, use 500,000+.

2. **Check gene set sizes**: Gene sets with fewer than 5 genes may yield unreliable results. Use `min.set.n` to filter.

3. **Monitor memory usage**: Large datasets with many gene sets can require substantial memory. Consider using fewer cores if memory is limited.

4. **Use appropriate coordinates**: If your data uses physical positions (bp), provide recombination rates for accurate distance calculations.

5. **Save intermediate results**: The pipeline can be run incrementally, so save objects after each step for troubleshooting.

### Common Pitfalls

1. **Mismatched IDs**: Ensure `objID` in `obj.info` matches `objID` in `set.obj`, and `setID` in `set.info` matches `setID` in `set.obj`.

2. **Missing coordinates**: Genomic coordinates are optional for `read_polylinkr_data()` but required for `rescale_polylinkr_data()`. If missing, rescaling will fail.

3. **Duplicate entries**: The functions handle duplicates automatically, but it's best to clean your data beforehand.

4. **Insufficient permutations**: Too few permutations lead to unstable p-values. The minimum recommended is 50,000.

5. **Forgetting the seed**: Always set a seed for reproducible results, especially when parallel processing.

### Interpreting Results

- **setScore.std**: Standardized gene set scores (mean = 0, variance = 1 per gene under null)
- **setScore.std.p**: P-values before autocorrelation correction
- **setScore.rs**: Rescaled scores accounting for genetic autocorrelation
- **setScore.rs.p**: P-values after autocorrelation correction (use these for final interpretation)

Gene sets with small p-values (< 0.05 or Bonferroni-corrected) and positive scores indicate enrichment for the selection signals in your input data.

## Session Information

```{r session-info}
sessionInfo()
```

## References

For more information about polylinkR:

- Package documentation: `?polylinkR`
- Function help: `?read_polylinkr_data`, `?permute_polylinkr_data`, `?rescale_polylinkr_data`, `?prune_polylinkr_data`
- GitHub repository: https://github.com/ACAD-UofA/PolyLinkR

For the underlying statistical methods:

- The circular permutation algorithm preserves linkage structure while generating null distributions
- Generalized Pareto Distribution fitting enables accurate estimation of small p-values
- Rescaling accounts for genetic autocorrelation using empirical variogram modeling
