---
title: "Controlling for Confounding Variables in Polygenic Linkage Analysis"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Controlling for Confounding Variables in Polygenic Linkage Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(polylinkR)
library(data.table)
```

## 1. Why Control for Confounders in Polygenic Linkage Analysis

Confounding variables are factors that influence both the genetic architecture and the trait being studied, potentially leading to spurious associations. In polygenic pathway enrichment analysis, common confounders include:

- **Age**: Gene expression and methylation patterns change with age
- **Sex**: X-chromosome inactivation and sex-specific genetic effects
- **Ancestry**: Population stratification can create LD patterns unrelated to trait biology

Without accounting for these factors, pathway enrichment tests may:

1. **Inflate false positive rates**: Genes correlated with confounders appear enriched
2. **Mask true signals**: Real pathway associations may be obscured by confounder-induced noise
3. **Produce irreproducible results**: Confounder-driven associations rarely replicate

polylinkR addresses this through a **local quadratic regression** approach that:

- Calculates Mahalanobis distances between genes in covariate space
- Applies kernel weighting to identify genetically proximal genes with similar covariate profiles
- Regresses out confounder effects while preserving biological signal

## 2. Creating Covariate Dataframes with Sample IDs Matching Genetic Data

The `Anatolia_EF_CLR` dataset contains gene-level statistics but no covariate columns. For this vignette, we'll create synthetic confounders that simulate real-world scenarios.

First, let's examine the structure of the data:

```{r examine-data}
# Load the example data
data(Anatolia_EF_CLR)

# Check the structure
str(Anatolia_EF_CLR)
head(Anatolia_EF_CLR)
```

Key columns for covariate integration:

- `objID`: Unique gene identifier (must match between obj.info and covariates)
- `objStat`: Gene-level test statistic (the score to be deconfounded)
- `chr`, `startpos`, `endpos`: Genomic coordinates (required for deconfounding)

Now, let's create synthetic covariates that simulate age, sex, and ancestry effects:

```{r create-covariates}
# Create a copy of the data for modification
obj_info <- copy(Anatolia_EF_CLR)

# Set seed for reproducibility
set.seed(42)
n_genes <- nrow(obj_info)

# Cov1: Age-related methylation proxy
# Genes on certain chromosomes tend to have age-related expression changes
obj_info[, Cov1 := rnorm(.N, mean = 0, sd = 1)]
# Add chromosome-specific patterns (simulating age effects)
obj_info[chr %in% c(1, 3, 5), Cov1 := Cov1 + 0.5]
obj_info[chr %in% c(2, 4, 6), Cov1 := Cov1 - 0.3]

# Cov2: Sex chromosome dosage
# X chromosome genes have different expression patterns
obj_info[, Cov2 := rnorm(.N, mean = 0, sd = 1)]
obj_info[chr == 23, Cov2 := Cov2 + 1.5]  # X chromosome enrichment
obj_info[chr == 24, Cov2 := Cov2 - 1.0]  # Y chromosome depletion

# Cov3: Ancestry principal component proxy
# Spatial autocorrelation simulating population structure
obj_info[, gene_mid := (startpos + endpos) / 2]
obj_info[, Cov3 := rnorm(.N, mean = 0, sd = 1)]
# Create regional patterns
for (chrom in unique(obj_info$chr)) {
  chr_genes <- which(obj_info$chr == chrom)
  if (length(chr_genes) > 10) {
    # Smooth spatial pattern
    positions <- obj_info$gene_mid[chr_genes]
    sorted_idx <- order(positions)
    trend <- sin(positions[sorted_idx] / max(positions) * 2 * pi)
    obj_info$Cov3[chr_genes[sorted_idx]] <- obj_info$Cov3[chr_genes[sorted_idx]] + trend
  }
}

# Remove temporary column
obj_info[, gene_mid := NULL]

# Display the new covariate columns
head(obj_info[, c("objID", "objName", "Cov1", "Cov2", "Cov3")])

# Summary statistics
summary(obj_info[, .(Cov1, Cov2, Cov3)])
```

Now let's simulate how these covariates might confound the gene scores:

```{r simulate-confounding}
# Add confounding effect to gene scores
# In real data, these effects are already present in the objStat values
# Here we demonstrate by adding synthetic confounding

obj_info[, objStat_original := objStat]

# Add covariate effects to simulate confounded data
obj_info[, objStat := objStat + 
           0.3 * Cov1 +    # Age effect
           0.2 * Cov2 +    # Sex effect  
           0.25 * Cov3]    # Ancestry effect

# Check correlation between covariates and scores
print("Correlations with confounded scores:")
cor_matrix <- cor(obj_info[, .(objStat, Cov1, Cov2, Cov3)])
print(cor_matrix)
```

## 3. Using the `covar` Parameter in `plR_permute()` and `plR_rescale()`

In polylinkR, covariates are detected automatically from columns with the `Cov` prefix in the `obj.info` file. The deconfounding process happens in `plR_permute()` through these key parameters:

- `md.meth`: Mahalanobis distance method (`"robust"` for rank-based, `"raw"` for Pearson)
- `kern.bound`: Genomic distance flanking region to exclude from kernel weights
- `kern.func`: Kernel function (`"normal"`, `"exponential"`, or `"inverse"`)
- `kern.scale`: Scaling parameter for kernel decay
- `kern.wt.max`: Maximum weight for any single gene

Let's prepare the polylinkR input with the covariate-enriched `obj.info`:

```{r prepare-input}
# Load additional required data
data(PolyLinkR_SetInfo)
data(PolyLinkR_SetObj)

# Create temporary directory for input files
temp_dir <- tempfile("polylinkR_covar_demo")
dir.create(temp_dir, recursive = TRUE)

# Write files with covariate columns
fwrite(obj_info, file.path(temp_dir, "ObjInfo.txt"), sep = "\t")
fwrite(PolyLinkR_SetInfo, file.path(temp_dir, "SetInfo.txt"), sep = "\t")
fwrite(PolyLinkR_SetObj, file.path(temp_dir, "SetObj.txt"), sep = "\t")

# Read the data into polylinkR
plr_obj <- plR_read(
  input.path = temp_dir,
  verbose = TRUE
)

# Verify the object structure
print(plr_obj)
```

## 4. Example with Synthetic Confounders on Anatolia_EF_CLR Data

### 4.1 Running Analysis WITH Covariate Control

```{r run-with-covar, eval=FALSE}
# Run deconfounding and permutation with covariate control
# Note: Using reduced permutations for demonstration (normally use 1e5 or more)

plr_with_covar <- plR_permute(
  plR.input = plr_obj,
  permute = TRUE,
  n.perm = 1e4,  # Reduced for demo; use 1e5+ for real analysis
  n.boot = 10,   # Reduced for demo
  md.meth = "robust",      # Use rank-based robust method
  kern.bound = 1e5,        # 100kb flanking region
  kern.func = "normal",    # Gaussian kernel
  kern.scale = 2,          # Standard decay
  kern.wt.max = 0.05,      # Max 5% weight for any gene
  seed = 42,
  verbose = TRUE,
  n.cores = 1
)
```

For this vignette, we'll demonstrate the setup:

```{r setup-covar-run}
# Check that covariates are properly detected
obj_info_check <- copy(plr_obj$obj.info)

# Identify covariate columns
cov_names <- grep("^Cov[0-9]+$", names(obj_info_check), value = TRUE)
cat("Detected covariate columns:", paste(cov_names, collapse = ", "), "\n")
cat("Number of covariates:", length(cov_names), "\n")
```

### 4.2 Running Analysis WITHOUT Covariate Control

To run without covariate control, simply exclude the `Cov` columns from `obj.info` or set `permute = FALSE` with covariates removed:

```{r run-without-covar}
# Create obj.info without covariate columns
obj_info_no_covar <- obj_info[, .(objID, objStat = objStat_original, objName, 
                                   GeneLength, chr, startpos, endpos)]

# Write to temporary directory
temp_dir2 <- tempfile("polylinkR_no_covar")
dir.create(temp_dir2, recursive = TRUE)

fwrite(obj_info_no_covar, file.path(temp_dir2, "ObjInfo.txt"), sep = "\t")
fwrite(PolyLinkR_SetInfo, file.path(temp_dir2, "SetInfo.txt"), sep = "\t")
fwrite(PolyLinkR_SetObj, file.path(temp_dir2, "SetObj.txt"), sep = "\t")

# Read without covariates
plr_no_covar <- plR_read(
  input.path = temp_dir2,
  verbose = TRUE
)

# Check that no covariates are detected
cat("Covariate columns in no-covar object:", 
    length(grep("^Cov[0-9]+$", names(plr_no_covar$obj.info))), "\n")
```

## 5. Interpreting Results With and Without Covariate Control

When comparing results with and without covariate control, look for these patterns:

### 5.1 Gene Score Changes

After deconfounding, examine how gene scores changed:

```{r compare-scores, eval=FALSE}
# Extract deconfounded scores (would be available after running plR_permute)
# deconfounded_scores <- plr_with_covar$obj.info$objStat_dc
# original_scores <- plr_obj$obj.info$objStat

# Plot comparison
# plot(original_scores, deconfounded_scores, 
#      xlab = "Original Scores", ylab = "Deconfounded Scores",
#      main = "Impact of Covariate Adjustment")
# abline(0, 1, col = "red", lty = 2)
```

### 5.2 Key Metrics to Compare

| Metric | Without Covariates | With Covariates | Interpretation |
|--------|-------------------|-----------------|----------------|
| Pathway p-values | May be inflated | Generally more conservative | Confounders can create artificial enrichment |
| Significant pathways | Potentially many false positives | Fewer but more reliable | Covariates reduce spurious associations |
| Gene score variance | Higher | Reduced (confounder component removed) | Deconfounding removes systematic variation |

### 5.3 Diagnostic Plots

When running with `verbose = TRUE`, polylinkR generates diagnostic information about:

1. **Covariate correlations**: How strongly each covariate correlates with gene scores
2. **Mahalanobis distances**: Distribution of distances in covariate space
3. **Kernel weights**: How genes are weighted in the local regression
4. **Deconfounding statistics**: Model fit and residual patterns

## 6. Best Practices for Selecting Covariates

### 6.1 What to Include

**Essential covariates:**
- Age (continuous or binned)
- Sex (binary or factor)
- Ancestry principal components (typically PC1-PC5)
- Technical factors (batch, platform, sequencing depth)

**Optional covariates:**
- Cell type proportions (for bulk tissue)
- BMI, smoking status (if relevant to trait)
- Sample collection time/season

### 6.2 What to Avoid

- **The trait of interest itself**: Never include the outcome variable as a covariate
- **Mediators**: Variables on the causal pathway between genes and trait
- **Colliders**: Variables influenced by both genes and trait
- **Too many PCs**: Including >10 ancestry PCs often captures noise rather than structure

### 6.3 Covariate Preprocessing

```{r preprocess-covariates}
# Example preprocessing pipeline
preprocess_covariates <- function(cov_data) {
  
  # 1. Handle missing values
  cov_data <- na.omit(cov_data)
  
  # 2. Standardize continuous covariates
  cont_cols <- c("Age", "BMI", "PC1", "PC2", "PC3")
  for (col in cont_cols) {
    if (col %in% names(cov_data)) {
      cov_data[[col]] <- scale(cov_data[[col]])[, 1]
    }
  }
  
  # 3. Encode categorical variables
  if ("Sex" %in% names(cov_data)) {
    cov_data$Sex <- as.numeric(factor(cov_data$Sex)) - 1  # 0/1 encoding
  }
  
  # 4. Remove collinear covariates
  cor_matrix <- cor(cov_data[, -1])  # Exclude ID column
  high_cor <- which(abs(cor_matrix) > 0.9 & upper.tri(cor_matrix), arr.ind = TRUE)
  if (nrow(high_cor) > 0) {
    cat("Highly correlated covariate pairs detected:\n")
    for (i in seq_len(nrow(high_cor))) {
      cat(names(cov_data)[high_cor[i, 1] + 1], "~", 
          names(cov_data)[high_cor[i, 2] + 1], "\n")
    }
  }
  
  return(cov_data)
}
```

### 6.4 Naming Convention

polylinkR automatically detects covariates using the pattern `Cov[0-9]+`. Ensure your covariate columns follow this naming scheme:

```{r naming-convention}
# Example: Proper naming for polylinkR
obj_info_named <- data.table(
  objID = 1:100,
  objStat = rnorm(100),
  Cov1 = rnorm(100),   # Age
  Cov2 = rbinom(100, 1, 0.5),  # Sex
  Cov3 = rnorm(100),   # PC1
  Cov4 = rnorm(100),   # PC2
  chr = sample(1:22, 100, replace = TRUE),
  startpos = sample(1e6:1e8, 100),
  endpos = sample(1e6:1e8, 100)
)

# Order doesn't matter, but the prefix must be exactly "Cov"
head(obj_info_named)
```

### 6.5 Parameter Tuning Guidelines

| Parameter | Default | When to Adjust | Conservative Setting |
|-----------|---------|----------------|---------------------|
| `md.meth` | `"robust"` | Use `"raw"` for normal-distributed covariates | `"robust"` (always recommended) |
| `kern.bound` | 1e5 | Increase for gene-dense regions | 2e5 or higher |
| `kern.scale` | 2 | Decrease for tighter local regression | 1 |
| `kern.wt.max` | 0.05 | Decrease to limit influence of single genes | 0.02 or 0.01 |

### 6.6 Validation Checklist

Before trusting deconfounded results:

- [ ] Covariate columns are correctly named (`Cov1`, `Cov2`, ...)
- [ ] No missing values in covariate columns
- [ ] Covariates correlate with raw gene scores (confounding exists)
- [ ] After deconfounding, gene score ~ covariate correlations are reduced
- [ ] Results are consistent across different random seeds
- [ ] Biological pathways of interest remain significant (or become more significant)

## Session Information

```{r session-info}
sessionInfo()

# Clean up temporary directories
unlink(temp_dir, recursive = TRUE)
unlink(temp_dir2, recursive = TRUE)
```

## References

1. **Population Stratification**: Price AL, et al. (2006) Principal components analysis corrects for stratification in genome-wide association studies. *Nature Genetics*.

2. **Batch Effects**: Leek JT, et al. (2010) Tackling the widespread and critical impact of batch effects in high-throughput data. *Nature Reviews Genetics*.

3. **Local Regression**: Cleveland WS (1979) Robust locally weighted regression and smoothing scatterplots. *Journal of the American Statistical Association*.

4. **Mahalanobis Distance**: Mahalanobis PC (1936) On the generalised distance in statistics. *Proceedings of the National Institute of Sciences of India*.
