---
title: "Input Formats and Parameters Reference"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Input Formats and Parameters Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Reference vignette with code examples
)
```

```{r setup}
library(polylinkR)
```

# Overview

This vignette serves as the comprehensive reference guide for **polylinkR** input formats and function parameters. Use this document as your go-to resource when preparing data or configuring analysis parameters.

---

# 1. Input Data Formats

## 1.1 Required Files

polylinkR requires three core input files that define genes, gene sets, and their relationships:

### ObjInfo File (Gene Information)

The gene information file contains one row per gene with the following columns:

| Column | Type | Required | Description |
|--------|------|----------|-------------|
| `objID` | character/numeric | Yes | Unique gene identifier |
| `objStat` | numeric | Conditional | Gene score/statistic (required if no `var.info`) |
| `chr` | character/numeric | Conditional | Chromosome/contig label (required for deconfounding/rescaling) |
| `startpos` | numeric | Conditional | Gene start position in base pairs |
| `endpos` | numeric | Conditional | Gene end position in base pairs |
| `startpos.base` | numeric | Conditional | Original gene start (without buffer) |
| `endpos.base` | numeric | Conditional | Original gene end (without buffer) |
| `Cov1`, `Cov2`, ... | numeric | No | Covariate columns for deconfounding |

**Example ObjInfo file:**

```{r objinfo-example, eval=FALSE}
# Example obj.info data structure
objID    chr   startpos   endpos    objStat   Cov1   Cov2
GENE001  1     1000000    1050000   2.34      0.5    1.2
GENE002  1     2050000    2100000   -1.87     0.3    0.8
GENE003  2     1500000    1600000   3.12      0.9    1.5
```

### SetInfo File (Gene Set Information)

The gene set information file contains one row per gene set:

| Column | Type | Required | Description |
|--------|------|----------|-------------|
| `setID` | character/numeric | Yes | Unique gene set identifier |

Optional columns (preserved but not required):
- `setName`: Human-readable set name
- `setSource`: Source database or annotation
- Any additional metadata columns

**Example SetInfo file:**

```{r setinfo-example, eval=FALSE}
# Example set.info data structure
setID      setName               setSource
SET001     Cell_cycle_genes      Reactome
SET002     DNA_repair_pathway    KEGG
SET003     Apoptosis_genes       GO_BP
```

### SetObj File (Gene-to-Set Mappings)

The mapping file defines which genes belong to which sets:

| Column | Type | Required | Description |
|--------|------|----------|-------------|
| `setID` | character/numeric | Yes | Gene set identifier (must match SetInfo) |
| `objID` | character/numeric | Yes | Gene identifier (must match ObjInfo) |

**Example SetObj file:**

```{r setobj-example, eval=FALSE}
# Example set.obj data structure
setID    objID
SET001   GENE001
SET001   GENE002
SET002   GENE001
SET002   GENE003
SET003   GENE002
SET003   GENE003
```

## 1.2 Optional Files

### VarInfo File (Variant/SNP Information)

Used to generate gene scores from variant-level data:

| Column | Type | Required | Description |
|--------|------|----------|-------------|
| `chr` | character/numeric | Yes | Chromosome/contig label |
| `pos` | numeric | Yes | Variant position |
| `value` | numeric | Yes | Statistical value (e.g., -log10(p), z-score) |

**Example VarInfo file:**

```{r varinfo-example, eval=FALSE}
# Example var.info data structure
chr   pos        value
1     1000023    3.45
1     1023456    2.78
1     2056789    4.12
2     1500123    1.98
```

### RecRate File (Recombination Rate)

Used to convert physical (bp) to genetic (cM) coordinates:

| Column | Type | Required | Description |
|--------|------|----------|-------------|
| `chr` | character/numeric | Yes | Chromosome label |
| `pos` | numeric | Yes | Position in base pairs |
| `rate` | numeric | Yes | Recombination rate (cM/Mb or cM/bp) |
| `map` | numeric | No | Genetic map position (cM) - will be calculated if absent |

**Format note:** Uses HapMap-style format. See examples at [Zenodo](https://zenodo.org/records/11437540).

## 1.3 File Format Flexibility

polylinkR accepts both comma-separated (.csv) and tab-separated (.tsv/.txt) files. Column names are case-insensitive and support multiple naming conventions:

- `objID`, `obj_ID`, `ObjID`, `obj.id`, `Obj_Id`
- `setID`, `set_ID`, `SetID`, `set.id`, `Set_Id`
- `startpos`, `start_pos`, `StartPos`, `start`
- `endpos`, `end_pos`, `EndPos`, `end`

## 1.4 Data Encoding

### Gene Score Encoding

Gene scores (`objStat`) can be encoded as:

- **Z-scores**: Standardized effect sizes (recommended)
- **-log10(p-values)**: Transformed p-values
- **CLR (Centered Log-Ratio)**: For compositional data

The encoding choice affects interpretation but the permutation framework is robust to the scale.

### Genotype Encoding (for variant data)

When using `var.info` to generate gene scores:

| Encoding | Description | Use Case |
|----------|-------------|----------|
| `-log10(p)` | Negative log-transformed p-values | Standard GWAS summary stats |
| Z-scores | Standardized effect sizes | Effect direction matters |
| Raw statistics | Original test statistics | Custom analyses |

---

# 2. Data Loading Reference

## 2.1 `plR_read()` Function

The `plR_read()` function is the entry point for loading and validating data.

### File Path Parameters

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `input_path` | character | `NULL` | Directory containing all input files |
| `object_info_path` | character | `NULL` | Direct path to obj.info file |
| `gene_set_info_path` | character | `NULL` | Direct path to set.info file |
| `gene_set_mapping_path` | character | `NULL` | Direct path to set.obj file |
| `var.info.path` | character | `NULL` | Direct path to var.info file |
| `rec.rate.path` | character | `NULL` | Direct path to rec.rate file |
| `group` | character | `NULL` | Label to identify files within directory |

### Filtering Parameters

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `min_set_size` | integer | `2` | Minimum gene set size to retain |
| `max_set_size` | numeric | `Inf` | Maximum gene set size to retain |
| `obj.in` | character/numeric | `NULL` | Gene IDs to explicitly retain |
| `obj.out` | character/numeric | `NULL` | Gene IDs to explicitly remove |
| `set.in` | character/numeric | `NULL` | Set IDs to explicitly retain |
| `set.out` | character/numeric | `NULL` | Set IDs to explicitly remove |
| `merge_threshold` | numeric | `0.95` | Minimum proportion of shared genes to merge sets |
| `rem.genes` | logical | `FALSE` | Remove genes with identical genomic positions |

### Gene Score Generation Parameters

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `obj.stat.fun` | character | `"non.param"` | Function for gene score correction: `"non.param"` or `"lm.logN"` |
| `obj.buffer` | numeric | `"auto"` | Buffer around genes (bp) for variant assignment |
| `bin.size` | integer | `250` | Bin size for non-parametric correction |

### Coordinate Conversion Parameters

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `map.fun` | character | `"kosambi"` | Mapping function: `"Haldane"`, `"Kosambi"`, `"Carter-Falconer"`, `"Morgan"` |

### General Parameters

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `verbose` | logical | `TRUE` | Print progress messages |

## 2.2 Validation Checks

`plR_read()` performs comprehensive validation:

1. **Column presence**: Ensures required columns exist with flexible naming
2. **Data completeness**: Removes rows with missing required values
3. **ID consistency**: Checks that IDs match across files
4. **Chromosome compatibility**: Validates chromosome labels in auxiliary files
5. **Duplicate removal**: Identifies and removes duplicate genes/sets
6. **Set merging**: Merges highly similar gene sets based on `merge_threshold`
7. **Size filtering**: Enforces `min_set_size` and `max_set_size` constraints

## 2.3 Loading Examples

### Example 1: Load from directory

```{r read-directory}
# Load all files from a single directory
plr_obj <- plR_read(
  input_path = "path/to/data",
  verbose = TRUE
)
```

### Example 2: Load with specific file paths

```{r read-paths}
# Specify individual file paths
plr_obj <- plR_read(
  object_info_path = "data/genes.tsv",
  gene_set_info_path = "data/sets.csv",
  gene_set_mapping_path = "data/mappings.txt",
  verbose = TRUE
)
```

### Example 3: Load with filtering

```{r read-filter}
# Load with gene set size constraints and merging
plr_obj <- plR_read(
  input_path = "path/to/data",
  min_set_size = 5,
  max_set_size = 500,
  merge_threshold = 0.85,
  rem.genes = TRUE
)
```

### Example 4: Generate gene scores from variant data

```{r read-varinfo}
# Generate gene scores from variant-level data
plr_obj <- plR_read(
  input_path = "path/to/data",
  var.info.path = "data/gwas_summary.tsv",
  obj.stat.fun = "non.param",
  obj.buffer = 50000,  # 50kb buffer
  bin.size = 250
)
```

### Example 5: Convert coordinates to genetic distance

```{r read-recrate}
# Convert physical to genetic coordinates
plr_obj <- plR_read(
  input_path = "path/to/data",
  rec.rate.path = "data/recombination.tsv",
  map.fun = "kosambi"
)
```

---

# 3. Function Parameters Reference Tables

## 3.1 `plR_permute()` Parameters

Gene set enrichment testing with confounder correction.

| Parameter | Type | Default | Range | Description |
|-----------|------|---------|-------|-------------|
| `plR.input` | plR object | Required | - | Input from `plR_read()` |
| `permute` | logical | `TRUE` | - | Perform permutation testing |
| `n.perm` | integer | `500000` | `[50000, Inf)` | Number of permutations (divisible by 10000) |
| `n.boot` | integer | `30` | `[5, Inf)` | Bootstrap replicates for null inference |
| `alt` | character | `"upper"` | `"upper"`, `"lower"` | Alternative hypothesis direction |
| `md.meth` | character | `"robust"` | `"robust"`, `"raw"` | Mahalanobis distance method |
| `kern.bound` | numeric | `"auto"` | `(0, Inf]` | Flanking region for gene exclusion |
| `kern.func` | character | `"normal"` | `"normal"`, `"exponential"`, `"inverse"` | Kernel function |
| `kern.scale` | numeric | `"auto"` | `(0, Inf]` | Kernel scaling parameter |
| `kern.wt.max` | numeric | `0.05` | `(1/(n-1), 1]` | Maximum probability weight |
| `gpd.cutoff` | numeric | `5000/n.perm` | `[max(1e-4, 500/n.perm), 0.05]` | GPD tail fitting threshold |
| `seed` | integer | `NULL` | `[-.Machine$integer.max, .Machine$integer.max]` | Random seed |
| `verbose` | logical | `TRUE` | - | Print progress messages |
| `n.cores` | integer/character | `"auto"` | `[1, max_cores]` or `"auto"` | Number of cores |
| `fut.plan` | character | `"auto"` | `"multisession"`, `"multicore"`, `"cluster"`, `"sequential"`, `"auto"` | Parallel backend |

### Parameter Details

**`n.perm`** - Higher values increase precision of small p-values but increase computation time. Default 500,000 provides good precision down to ~1e-5.

**`n.boot`** - More bootstrap replicates improve estimation of the null distribution quantiles.

**`alt`** - Use `"upper"` for testing enrichment in genes with high scores (e.g., positive selection), `"lower"` for low scores (e.g., purifying selection).

**`md.meth`** - `"robust"` converts covariates to ranks before computing Mahalanobis distances (recommended for non-normal covariates). `"raw"` uses original values.

**`kern.bound`** - Defines a buffer region around each gene where overlapping genes receive reduced weights. `"auto"` sets 0.1 Mbp or 0.1 cM.

**`kern.func`** - Determines how distance-to-weight conversion occurs:
- `"normal"`: Gaussian decay
- `"exponential"`: Exponential decay
- `"inverse"`: Inverse distance weighting

## 3.2 `plR_rescale()` Parameters

Rescale gene set scores for genetic autocorrelation.

| Parameter | Type | Default | Range | Description |
|-----------|------|---------|-------|-------------|
| `plR.input` | plR object | Required | - | Input from `plR_permute()` |
| `rescale` | logical | `TRUE` | - | Perform rescaling step |
| `fast` | logical | `TRUE` | - | Use fast analytical rescaling |
| `ac` | data.table/data.frame | `NULL` | - | User-provided autocovariance object |
| `cgm.range` | numeric/character | `"auto"` | `[1e5, 5e7]` bp or `[0.1, 50]` cM | Maximum inter-gene lag for autocovariance |
| `cgm.bin` | numeric | `30` | `[10, 1000]` | Minimum gene pairs per lag bin |
| `cgm.wt.max` | numeric | `0.05` | `[2/(nBins-1), 1]` | Maximum probability weight for lag windows |
| `emp.bayes` | character | `"auto"` | `"auto"`, `"full"`, `"reduced"` | Empirical Bayes framework |
| `min.rho` | numeric | `1e-5` | `(0, 0.01]` | Minimum correlation threshold |
| `verbose` | logical | `TRUE` | - | Print progress messages |
| `n.cores` | integer/character | `"auto"` | `[1, max_cores]` or `"auto"` | Number of cores |
| `fut.plan` | character | `"auto"` | `"multisession"`, `"multicore"`, `"cluster"`, `"sequential"`, `"auto"` | Parallel backend |

### Parameter Details

**`fast`** - When `TRUE`, uses analytical rescaling (much faster). When `FALSE`, performs full empirical rescaling (more accurate but computationally intensive).

**`cgm.range`** - Defines the maximum distance for computing empirical autocovariance. `"auto"` sets 3 Mbp or 3 cM.

**`cgm.bin`** - Controls binning of empirical covariances. Larger values create smoother correlograms but may miss local structure.

**`emp.bayes`** - Framework for regularizing chromosome-level parameters:
- `"full"`: Correlated random effects (requires ≥15 chromosomes)
- `"reduced"`: Independent random effects
- `"auto"`: Chooses based on chromosome count

**`min.rho`** - Correlations below this threshold are set to zero, reducing noise in the covariance matrix.

## 3.3 `plR_prune()` Parameters

Prune gene sets and compute FDR-corrected p-values.

| Parameter | Type | Default | Range | Description |
|-----------|------|---------|-------|-------------|
| `plR.input` | plR object | Required | - | Input from `plR_permute()` or `plR_rescale()` |
| `n.fdr` | integer | `300` | `[100, Inf)`, divisible by 100 | Pruning replicates for FDR |
| `est.pi0` | logical | `TRUE` | - | Estimate proportion of true nulls |
| `tolerance` | numeric | `0.001` | `(0, 0.01]` | Convergence tolerance for pi0 |
| `verbose` | logical | `TRUE` | - | Print progress messages |
| `n.cores` | integer/character | `"auto"` | `[1, max_cores]` or `"auto"` | Number of cores |
| `fut.plan` | character | `"auto"` | `"multisession"`, `"multicore"`, `"cluster"`, `"sequential"`, `"auto"` | Parallel backend |

### Parameter Details

**`n.fdr`** - Number of permuted gene set replicates used to estimate the null distribution for FDR. More replicates improve q-value precision.

**`est.pi0`** - When `TRUE`, estimates the proportion of true null hypotheses (pi0). When `FALSE`, assumes pi0 = 1 (conservative).

**`tolerance`** - Controls when pi0 estimation stops. Smaller values require more iterations for convergence.

## 3.4 Distance Calculation Methods

### Mahalanobis Distance (`md.meth`)

Used in confounder correction (`plR_permute()`):

| Method | Description | When to Use |
|--------|-------------|-------------|
| `"robust"` | Ranks converted to normal scores, Spearman covariance | Non-normal covariates, outliers present |
| `"raw"` | Original values, Pearson covariance | Normal covariates, no extreme outliers |

### Distance Metrics (implicit)

For genetic coordinate systems:

| System | Units | Use Case |
|--------|-------|----------|
| Physical | Base pairs (bp) | Fine-scale analyses, species without genetic maps |
| Genetic | CentiMorgans (cM) | Cross-species comparisons, LD-aware analyses |

---

# 4. Output Format Specifications

## 4.1 plR Object Structure

All polylinkR functions return S3 objects of class `plR` containing three data.tables:

### `obj.info` - Gene Information

| Column | Description |
|--------|-------------|
| `objID` | Gene identifier |
| `objID.orig` | Original gene identifier |
| `chr` | Chromosome |
| `startpos` | Start position (including buffer) |
| `endpos` | End position (including buffer) |
| `startpos.base` | Original start position |
| `endpos.base` | Original end position |
| `midpos` | Midpoint position |
| `objStat` | Original gene score |
| `objStat.res` | Residual gene score (after prognostic correction) |
| `objStat.std` | Standardized gene score (mean=0, sd=1) |
| `Cov1`, `Cov2`, ... | Covariate values (if provided) |

### `set.info` - Gene Set Information

| Column | Description |
|--------|-------------|
| `setID` | Gene set identifier |
| `setID.orig` | Original set identifier |
| `setN` | Number of genes in set |
| `setScore.std` | Standardized set score (from `plR_permute()`) |
| `setScore.std.p` | P-value for standardized score |
| `setScore.rs` | Rescaled set score (from `plR_rescale()`) |
| `setScore.rs.p` | P-value for rescaled score |
| `setScore.pr` | Pruned set score (from `plR_prune()`) |
| `setScore.pr.p` | Pruned p-value |
| `setScore.pr.q` | FDR-corrected q-value |
| `Rank` | Pruning rank |
| `Tot.obj.rem` | Total genes removed during pruning |
| `Tot.set.rem` | Total sets removed during pruning |

### `set.obj` - Gene-to-Set Mappings

| Column | Description |
|--------|-------------|
| `setID` | Gene set identifier |
| `objID` | Gene identifier |

## 4.2 Object Attributes

Access using `attributes(plr_obj)$attribute_name`:

| Attribute | Description | Access Example |
|-----------|-------------|----------------|
| `plR.data` | Reusable datasets and parameters | `attributes(plr)$plR.data$permute.data` |
| `plR.args` | Argument settings used | `attributes(plr)$plR.args$read.args` |
| `plR.summary` | Summary statistics | `attributes(plr)$plR.summary$permute.summary` |
| `plR.session` | R session info and run times | `attributes(plr)$plR.session$read.session` |
| `plR.seed` | Random seeds used | `attributes(plr)$plR.seed$seed` |
| `class` | S3 class identifier | Always `"plR"` |

## 4.3 Results Interpretation

### P-value Columns

| Column | When Significant | Interpretation |
|--------|------------------|----------------|
| `setScore.std.p` | < 0.05 | Gene set shows enrichment before LD correction |
| `setScore.rs.p` | < 0.05 | Gene set shows enrichment after LD correction |
| `setScore.pr.p` | < 0.05 | Gene set significant after pruning |
| `setScore.pr.q` | < 0.05 | Gene set significant after FDR correction |

### Score Columns

| Column | Meaning |
|--------|---------|
| `setScore.std` | Mean of standardized gene scores × sqrt(set size) |
| `setScore.rs` | Rescaled score accounting for gene correlations |
| `setScore.pr` | Score after removing shared genes with higher-ranked sets |

**Important:** For final interpretation, use `setScore.pr.q` (FDR-corrected q-values from the pruning step). This accounts for both LD structure and shared genes between pathways.

---

# 5. Troubleshooting Common Input Issues

## 5.1 Mismatched Sample IDs

### Problem
Gene IDs in `set.obj` don't match those in `obj.info`.

### Error Message
```
Some set.obj genes [objID] were not found in obj.info
```

### Solutions

1. **Check ID formats**: Ensure consistent naming (case-sensitive):
   ```{r check-ids}
   # Check overlap between files
   obj_ids <- obj.info$objID
   set_obj_ids <- unique(set.obj$objID)
   
   # Find mismatches
   missing_in_obj <- setdiff(set_obj_ids, obj_ids)
   missing_in_set <- setdiff(obj_ids, set_obj_ids)
   
   print(missing_in_obj)
   ```

2. **Standardize IDs**: Convert all IDs to the same case/format:
   ```{r standardize-ids}
   # Make all IDs uppercase
   obj.info$objID <- toupper(obj.info$objID)
   set.obj$objID <- toupper(set.obj$objID)
   ```

3. **Use ID mapping**: Create a translation table if IDs differ systematically.

## 5.2 Missing Data Handling

### Problem
Rows with missing values are being removed.

### Warning Message
```
5 incomplete rows detected and removed from obj.info
```

### Solutions

1. **Identify missing data**:
   ```{r find-missing}
   # Find rows with missing values
   incomplete <- !complete.cases(obj.info)
   print(obj.info[incomplete, ])
   ```

2. **Impute or fill**: For optional columns, add placeholder values:
   ```{r fill-missing}
   # Fill missing covariates with 0
   obj.info$Cov1[is.na(obj.info$Cov1)] <- 0
   ```

3. **Check file format**: Ensure proper delimiter handling:
   ```{r check-delimiter}
   # For tab-separated files
   obj.info <- data.table::fread("data.tsv", sep = "\t")
   ```

## 5.3 Memory Errors with Large Datasets

### Problem
R runs out of memory during analysis.

### Error Message
```
Error: cannot allocate vector of size ...
```

### Solutions

1. **Reduce permutations**: Start with fewer permutations:
   ```{r reduce-perm}
   plr_perm <- plR_permute(
     plR.input = plr_obj,
     n.perm = 100000,  # Reduced from 500000
     n.boot = 10       # Reduced from 30
   )
   ```

2. **Use fewer cores**: Reduce parallel overhead:
   ```{r reduce-cores}
   plr_perm <- plR_permute(
     plR.input = plr_obj,
     n.cores = 2  # Limit parallel cores
   )
   ```

3. **Process chromosomes separately**: Split by chromosome if appropriate.

4. **Increase system memory**: Use a machine with more RAM or use memory-mapped files.

## 5.4 Format Conversion Examples

### Converting from PLINK/association output

```{r convert-plink}
# PLINK .assoc output to var.info format
plink_output <- read.table("plink.assoc", header = TRUE)

var.info <- data.frame(
  chr = plink_output$CHR,
  pos = plink_output$BP,
  value = -log10(plink_output$P)  # Convert p-value to -log10
)

write.table(var.info, "var_info.tsv", sep = "\t", row.names = FALSE)
```

### Converting from VCF/GWAS catalog

```{r convert-gwas}
# Standard GWAS summary statistics
# Columns: SNP, CHR, BP, P, BETA, SE

gwas <- read.table("gwas_summary.tsv", header = TRUE)

# Option 1: Use -log10 p-values
var.info <- data.frame(
  chr = gwas$CHR,
  pos = gwas$BP,
  value = -log10(gwas$P)
)

# Option 2: Use Z-scores (if BETA and SE available)
var.info <- data.frame(
  chr = gwas$CHR,
  pos = gwas$BP,
  value = gwas$BETA / gwas$SE
)
```

### Converting gene annotation formats

```{r convert-genes}
# From GTF/GFF to obj.info
# Example using rtracklayer (not run)
# library(rtracklayer)
# gtf <- import("genes.gtf")
# 
# obj.info <- data.frame(
#   objID = gtf$gene_id,
#   chr = seqnames(gtf),
#   startpos = start(gtf),
#   endpos = end(gtf)
# )
```

### Converting from MSigDB/other pathway databases

```{r convert-pathways}
# From MSigDB GMT format to polylinkR format

# Read GMT file
gmt_lines <- readLines("h.all.v7.2.symbols.gmt")

# Parse to set.info and set.obj
set_info_list <- list()
set_obj_list <- list()

for (line in gmt_lines) {
  parts <- strsplit(line, "\t")[[1]]
  set_id <- parts[1]
  set_desc <- parts[2]
  genes <- parts[-(1:2)]
  
  # Add to set.info
  set_info_list[[length(set_info_list) + 1]] <- data.frame(
    setID = set_id,
    setName = set_desc
  )
  
  # Add to set.obj
  set_obj_list[[length(set_obj_list) + 1]] <- data.frame(
    setID = set_id,
    objID = genes
  )
}

set.info <- do.call(rbind, set_info_list)
set.obj <- do.call(rbind, set_obj_list)
```

## 5.5 Common Warnings and Their Meanings

| Warning | Meaning | Action |
|---------|---------|--------|
| "GPD fitting failed for >50% of nulls" | Extreme value distribution fitting unsuccessful | Try different `gpd.cutoff` or reduce `n.perm` |
| "Some obj.info chromosomes not present in rec.rate" | Missing recombination data for some chromosomes | Check chromosome naming consistency |
| "Short chromosome lengths detected" | Coordinates may already be in cM | Verify coordinate system |
| "Covariate matrix is singular" | Redundant covariates detected | Remove correlated covariates or reduce number |
| "Correlogram function fitting did not converge" | Autocovariance model fitting failed | Adjust `cgm.bin` or `cgm.range` |

---

# 6. Quick Reference Card

## One-Page Parameter Summary

### Essential Parameters by Analysis Stage

#### Stage 1: Data Loading (`plR_read()`)

| Parameter | Default | When to Change |
|-----------|---------|----------------|
| `min_set_size` | 2 | Increase to focus on larger pathways (e.g., 10 for robustness) |
| `max_set_size` | Inf | Decrease to exclude very large pathways (e.g., 1000) |
| `merge_threshold` | 0.95 | Decrease to merge more similar sets (e.g., 0.85) |
| `rem.genes` | FALSE | Set TRUE if duplicate gene positions expected |
| `obj.buffer` | auto | Increase for broader variant capture (e.g., 50e3 for 50kb) |

#### Stage 2: Permutation (`plR_permute()`)

| Parameter | Default | When to Change |
|-----------|---------|----------------|
| `n.perm` | 500000 | Increase for more precise small p-values (e.g., 1e6) |
| `n.boot` | 30 | Increase for smoother null distribution (e.g., 100) |
| `alt` | "upper" | Use "lower" for testing depletion |
| `md.meth` | "robust" | Use "raw" if covariates are normally distributed |
| `kern.bound` | auto | Increase to exclude larger regions (e.g., 1e6 for 1Mbp) |
| `kern.wt.max` | 0.05 | Increase if single genes should have more influence |
| `gpd.cutoff` | 5000/n.perm | Decrease for more conservative tail estimation |
| `n.cores` | auto | Set to specific value for reproducible parallelization |

#### Stage 3: Rescaling (`plR_rescale()`)

| Parameter | Default | When to Change |
|-----------|---------|----------------|
| `rescale` | TRUE | Set FALSE to skip (compute autocovariance only) |
| `fast` | TRUE | Set FALSE for empirical rescaling (slower, more accurate) |
| `cgm.range` | auto | Increase for longer-range LD (e.g., 5e6 for 5Mbp) |
| `cgm.bin` | 30 | Increase for smoother correlogram (e.g., 100) |
| `emp.bayes` | auto | Use "reduced" for few chromosomes (<15) |
| `min.rho` | 1e-5 | Increase to filter weak correlations (e.g., 1e-4) |

#### Stage 4: Pruning (`plR_prune()`)

| Parameter | Default | When to Change |
|-----------|---------|----------------|
| `n.fdr` | 300 | Increase for more precise q-values (e.g., 1000) |
| `est.pi0` | TRUE | Set FALSE for conservative q-values (pi0 = 1) |
| `tolerance` | 0.001 | Decrease for stricter pi0 convergence |

## Recommended Defaults by Analysis Type

### Quick GWAS Pathway Analysis

For a standard human GWAS with ~20,000 genes and ~100 pathways:

```{r quick-gwas, eval=FALSE}
# Load data
plr <- plR_read(
  input_path = "path/to/gwas_data",
  min_set_size = 10,
  merge_threshold = 0.90,
  verbose = TRUE
)

# Quick permutation (reduced precision)
plr <- plR_permute(
  plR.input = plr,
  n.perm = 100000,
  n.boot = 10,
  n.cores = 4
)

# Rescale for LD
plr <- plR_rescale(
  plR.input = plr,
  fast = TRUE,
  n.cores = 4
)

# Prune and get q-values
plr <- plR_prune(
  plR.input = plr,
  n.fdr = 300,
  n.cores = 4
)
```

### High-Precision Discovery Analysis

For publication-quality results with precise small p-values:

```{r high-precision, eval=FALSE}
# Load with stringent filtering
plr <- plR_read(
  input_path = "path/to/data",
  min_set_size = 5,
  max_set_size = 500,
  merge_threshold = 0.95,
  rem.genes = TRUE
)

# Extensive permutations
plr <- plR_permute(
  plR.input = plr,
  n.perm = 1000000,   # 1 million permutations
  n.boot = 100,       # 100 bootstrap replicates
  gpd.cutoff = 0.001  # Conservative GPD threshold
)

# Empirical rescaling
plr <- plR_rescale(
  plR.input = plr,
  fast = FALSE,       # Full empirical rescaling
  cgm.bin = 100       # Finer correlogram bins
)

# Extensive FDR
plr <- plR_prune(
  plR.input = plr,
  n.fdr = 1000        # Many replicates for precise q-values
)
```

### Non-Human Species / No LD Map

For species without recombination maps or when physical distance is preferred:

```{r no-ldmap, eval=FALSE}
# Load without coordinate conversion
# Use physical distances (bp) only
plr <- plR_read(
  input_path = "path/to/data",
  verbose = TRUE
)

# Standard permutation
plr <- plR_permute(plR.input = plr)

# Rescale with physical distance
# cgm.range will be in base pairs
plr <- plR_rescale(
  plR.input = plr,
  cgm.range = 2e6  # 2 Mbp range
)

plr <- plR_prune(plR.input = plr)
```

### Covariate-Aware Analysis

When including confounders (e.g., gene length, GC content):

```{r covariates, eval=FALSE}
# Ensure covariate columns are in obj.info
# obj.info should contain: Cov1, Cov2, etc.

# Load data
plr <- plR_read(input_path = "path/to/data")

# Permutation with robust covariate handling
plr <- plR_permute(
  plR.input = plr,
  md.meth = "robust",      # Robust Mahalanobis distances
  kern.bound = 1e6,        # 1 Mbp exclusion zone
  kern.wt.max = 0.10,      # Allow higher single-gene weights
  kern.func = "normal"     # Gaussian kernel
)

# Continue with rescale and prune
plr <- plR_rescale(plR.input = plr)
plr <- plR_prune(plR.input = plr)
```

## Parameter Cross-Reference

| Goal | Parameter | Function |
|------|-----------|----------|
| More precise p-values | `n.perm`, `n.boot` | `plR_permute()` |
| Conservative tail fitting | `gpd.cutoff` | `plR_permute()` |
| Better covariate handling | `md.meth`, `kern.*` | `plR_permute()` |
| Faster computation | `fast = TRUE` | `plR_rescale()` |
| Better LD correction | `fast = FALSE`, `cgm.*` | `plR_rescale()` |
| Precise q-values | `n.fdr` | `plR_prune()` |
| Conservative FDR | `est.pi0 = FALSE` | `plR_prune()` |
| Faster parallelization | `n.cores`, `fut.plan` | All functions |

## File Format Checklist

Before running analysis, verify:

- [ ] `objID` values match between `obj.info` and `set.obj`
- [ ] `setID` values match between `set.info` and `set.obj`
- [ ] Chromosome labels are consistent across all files
- [ ] Position columns are numeric (not factor/character)
- [ ] Gene scores (`objStat`) are present or `var.info` provided
- [ ] Covariate columns follow `Cov1`, `Cov2`, ... naming
- [ ] File delimiters are consistent (.csv or .tsv)
- [ ] No trailing whitespace in ID columns
- [ ] No duplicate `objID` or `setID` rows

## Output Column Quick Guide

| Result Type | Column Name | From Function |
|-------------|-------------|---------------|
| Raw set scores | `setScore.std` | `plR_permute()` |
| Raw p-values | `setScore.std.p` | `plR_permute()` |
| LD-corrected scores | `setScore.rs` | `plR_rescale()` |
| LD-corrected p-values | `setScore.rs.p` | `plR_rescale()` |
| Pruned scores | `setScore.pr` | `plR_prune()` |
| Pruned p-values | `setScore.pr.p` | `plR_prune()` |
| Final q-values | `setScore.pr.q` | `plR_prune()` |

**For final results, always use `setScore.pr.q` (FDR-corrected q-values).**

---

# Cross-Reference with Other Vignettes

- **Getting Started with polylinkR**: Basic workflow and runnable examples
- **Function Documentation**: `?plR_read`, `?plR_permute`, `?plR_rescale`, `?plR_prune`

---

# Session Information

```{r session-info, eval=TRUE}
sessionInfo()
```
