---
title: "Getting Started with polylinkR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with polylinkR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE  # This vignette contains runnable examples
)
```

```{r setup}
library(polylinkR)
```

## Overview

**polylinkR** is an R package for gene-based pathway enrichment that accounts for linkage disequilibrium among loci. This vignette demonstrates the basic workflow using the included `tiny_polylinkR` example dataset.

## The Four-Step Workflow

The polylinkR workflow follows four sequential steps:

1. **plR_read()** - Read and validate input files
2. **plR_permute()** - Generate permutation nulls and compute p-values
3. **plR_rescale()** - Rescale for genetic autocorrelation
4. **plR_prune()** - Prune gene sets and compute FDR

## Step 1: Read Input Data

First, we read the example data included with the package:

```{r read-data}
# Locate the example data
data_path <- system.file("extdata", "tiny_polylinkR", package = "polylinkR")
data_path

# Read the data
plr_obj <- plR_read(
  input.path = data_path,
  verbose = FALSE
)

# Check the result
class(plr_obj)
print(plr_obj)
```

The `plR_read()` function validates the input files and creates a `plR` class object containing:

- **set.info**: Gene set information
- **obj.info**: Gene (object) information with scores
- **set.obj**: Gene-to-set mappings

## Inspecting the plR Object

You can inspect the plR object using standard methods:

```{r inspect}
# Summary of the object
summary(plr_obj)

# Structure of the object
str(plr_obj)
```

## Input File Formats

polylinkR requires three input files:

### 1. ObjInfo.txt (Gene Information)

Contains gene-level information:

- **objID**: Unique gene identifier
- **objStat**: Gene-level statistic (e.g., p-value, z-score)
- Optional columns: chr, startpos, endpos (for coordinate-based analysis)

### 2. SetInfo.txt (Gene Set Information)

Contains gene set definitions:

- **setID**: Unique set identifier
- **setName**: Human-readable set name
- **setSource**: Source database (optional)

### 3. SetObj.txt (Gene-Set Mappings)

Contains gene-to-set relationships:

- **setID**: Set identifier (must match SetInfo.txt)
- **objID**: Gene identifier (must match ObjInfo.txt)

## Important Notes

### Active Development

This package is under active development. Results should be independently validated before use in production or clinical decisions.

### Algorithm Constraints

The core statistical algorithms (permutation nulls, GPD fitting, rescaling, pruning) are frozen and should not be modified without explicit maintainer approval.

## Next Steps

After reading the data, you would typically proceed through the pipeline:

1. **plR_permute()** - Generate permutation-based null distributions
2. **plR_rescale()** - Account for genetic autocorrelation
3. **plR_prune()** - Identify significant gene sets

See the function documentation (`?plR_permute`, `?plR_rescale`, `?plR_prune`) for details on subsequent steps.

## Session Information

```{r session-info}
sessionInfo()
```
