---
title: "Getting started with polylinkR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with polylinkR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

**polylinkR** implements gene-set analyses that respect linkage disequilibrium amongst markers by building a permutation null that preserves linkage structure. The public workflow is sequential: `plR_read()`, `plR_permute()`, `plR_rescale()`, and `plR_prune()`.

This package is **under active development**. Treat outputs as exploratory until you have reproduced them and validated them against independent evidence.

## Inputs

You supply tabular inputs describing objects (typically genes), gene sets, and membership. See `?plR_read` for required columns, optional covariate and recombination-rate files, and filtering options. Small example tables ship in `inst/extdata/` (when present) or `data/` for local experimentation.

## Next steps

```{r, eval = FALSE}
library(polylinkR)
# plR_read(input.path = "path/to/tables", verbose = TRUE)
```

Consult the function reference (`plR_read`, `plR_permute`, `plR_rescale`, `plR_prune`) for argument details, parallel execution via **future**, and random-seed behaviour.
